---
title: "Analysis of Vance County EMS Ambulance Distribution"
format: pdf
editor: visual
---

## 1 Background

The Vance County Emergency Medical Services system currently operates four ambulances and two stations with one located in the Southern district and one in the Central district. This configuration led to the residents in the Northern district being under-served with higher average response times than the other two districts. Such delays can lead to critical differences in patient outcomes especially in life threatening emergencies.

In response to this issue and demand for EMS coverage in the North region, Vance county is evaluating the potential of adding a new station in one of the two northern locations: near north or far north station. We use historical EMS call data to make a decision. Each record in the dataset represents one individual emergency trip which includes information like dispatch station, patient location coordinates, and time logs for dispatch, arrival, hospital transport, and clearance.

The main motivation of this analysis is to assess how travel times and system load vary across different station allocation scenarios which are listed in table 1 of the appendix and in particular we aim to answer three central questions: 1) which of the two possible Northern station locations would better serve the community through faster response time 2) which relocation (South or Central) would yield to improved response time and 3) how should the four available ambulances be allocated across stations to optimize service coverage and response times.

## 2 Exploratory Data Analysis

We modeled the distribution of best-guess travel time and from the histogram shown in figure 1 we can see that south station has a shorter median travel time and both the near north and far north stations have a longer travel time with a wider variability. We can see that the far north has the widest spread and the most outliers which implies more inconsistent travel times possibly stemming from it being so far up north by itself.

Talk about load conflict

mixed model thing

## 3 Modeling

Before the modeling we make some assumptions on how we will dispatch the ambulances. For scenario 0 which is our baseline where we just have 3 ambulances in central and 1 in the south station, we will assume that all north and central calls are taken by the central station and the south calls are taken by the south station ambulances. For scenarios 1 and 2 which are still 3 ambulances in the central but we move the south station to either the near north or far north station respectively, we assume that north calls are taken only by the North station and the central and south calls are only taken by the central station ambulances. Finally for scenarios 3 and 4 which are the ones where we keep the 1 south ambulance in the south station but move one of the central stations to either the near north or far north station respectively, we just assume all calls are taken by the same region station ambulance. The only exception to our modeling assumptions is if there are 3+ load conflicts at a assigned station, we make the available ambulance from the next closest station to be dispatched to take the call.

Under the assumption of a hierarchical approach where we condition on the calls where there is a total change in time, we fit a couple of models. The covariates are _______. They are both linear mixed models where utilize a random effect for a call since there is within group variation for each call. One takes into account heteroscedasticity concern and varies potential errors by longitude and latitude. 




## 4 Results

We note that there are different residuals for the aforementioned fits. We find that when examining the normality condition, that the model without the error variations seems to be a better fit for the data as seen in (ref the Q-Q plots), and consequently we analyze the results under the fit without the additional variation. 


## 5 Conclusion and Future Work

Our analysis showed that relocation to far north resulted in faster response times than near north. We have also shown that it is better to move central ambulances up north than moving the south ambulance and saw that relocating one ambulance from central to far north station reduces response times by about 4%. Therefore from our analysis we saw scenario 4 which is 1 far north, 2 central, and 1 south was the most optimal layout of the 4 ambulances as it reduces the average response time by 4% compared to the baseline of scenario 0.

Some limitations is that our model assumes traffic level to be constant throughout non-vs rush hour. We also are dealing with data that has error introduced by data randomization from the HIPAA protection so there is some marginal error in the distance calculation that stems from this. We are also limited from the smaller sample size and also from the best guess Google API which assumes average conditions.

In the future we can try adding time of day for a random effects which could provide a deeper insight into temporal variation in the response times. We can model this using a continous smooth spline approach using mgcr or poly to capture the nonlinear daily trends without overfitting. We can also try a Bayesian approach to improve interpretability, particularly in regions with fewer calls. Through partial pooling and uncertainty propagation, the use of this approach would allow information sharing across districts while accounting for data sparsity. Finally, a hierarchical modeling approach could be implemented to account for further subdivisions in call type and urgency and this would allow our model to distinguish between the emergency severity levels and incorporate medical criteria for the urgency of each call which could change how we allocate the ambulances across the systems.

## 6 Appendix


```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(lme4)
library(IRanges)
library(broom.mixed)
library(tidyverse)
library(nlme)
library(sjPlot)
if (!requireNamespace("IRanges", quietly = TRUE)) {
  install.packages("BiocManager")
  BiocManager::install("IRanges")
}
load("emsData.RData")
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
x<- x %>%
  mutate(
    disp_hour = hour(DT.DISP),
    disp_min  = minute(DT.DISP),
    rush_hour = !is.na(DT.DISP) & (
      
      #morning 7:30pm - 9:30am
      (disp_hour == 7 & disp_min >= 30) |
      (disp_hour == 8) |
      (disp_hour == 9 & disp_min < 30) |
      
      #evening 4:00pm - 6:30pm
      (disp_hour == 16) |
      (disp_hour == 17) |
      (disp_hour == 18 & disp_min < 30)
    ),
    # 1 if rush hour 0 else
    rush_hour_ind = as.integer(rush_hour)  
  ) %>%
  select(-disp_hour, -disp_min)

x_expanded <- x %>%
  # create a unique ID per original row
  mutate(CallID = row_number()) %>%
  
  # repeat rows for 5 scenarios
  tidyr::uncount(weights = 5, .id = "scenario_id") %>%
  mutate(
    Scenario = paste0("S", scenario_id - 1),
    
    # extract call location from REF.GRID
    CallLoc = str_extract(REF.GRID, "(South|Central|North)"),
    # normalize to codes
    CallLocCode = case_when(
      CallLoc == "South"   ~ "So",
      CallLoc == "Central" ~ "Ce",
      CallLoc == "North"   ~ "NN"  # treat all "North" as "NN"
    ),
    
    # dispatch rules
    Dispatch = case_when(
      # S0
      Scenario == "S0" & CallLocCode == "NN" ~ "Ce",
      Scenario == "S0" & CallLocCode == "Ce" ~ "Ce",
      Scenario == "S0" & CallLocCode == "So" ~ "So",
      
      # S1
      Scenario == "S1" & CallLocCode == "NN" ~ "NN",
      Scenario == "S1" & CallLocCode == "Ce" ~ "Ce",
      Scenario == "S1" & CallLocCode == "So" ~ "Ce",
      
      # S2
      Scenario == "S2" & CallLocCode == "NN" ~ "FN",
      Scenario == "S2" & CallLocCode == "Ce" ~ "Ce",
      Scenario == "S2" & CallLocCode == "So" ~ "Ce",
      
      # S3
      Scenario == "S3" & CallLocCode == "NN" ~ "NN",
      Scenario == "S3" & CallLocCode == "Ce" ~ "Ce",
      Scenario == "S3" & CallLocCode == "So" ~ "So",
      
      # S4
      Scenario == "S4" & CallLocCode == "NN" ~ "FN",
      Scenario == "S4" & CallLocCode == "Ce" ~ "Ce",
      Scenario == "S4" & CallLocCode == "So" ~ "So"
    ),
    
    # pick the correct estimated travel time
    EstTravelTime = case_when(
      Dispatch == "So" ~ eTT.BG.So,
      Dispatch == "Ce" ~ eTT.BG.Ce,
      Dispatch == "NN" ~ eTT.BG.NN,
      Dispatch == "FN" ~ eTT.BG.FN
    )
  ) %>%
  select(-scenario_id)

x_expanded <- x_expanded %>%
  mutate(Distance = case_when(
    Dispatch == "So" ~ Dist.So,
    Dispatch == "Ce" ~ Dist.Ce,
    Dispatch == "NN" ~ Dist.NN,
    Dispatch == "FN" ~ Dist.FN
  ))

```


## Exploratory Data Analysis

```{r, echo = FALSE, warning = FALSE,message=FALSE}

#don't remove central, fix this.
#we are assuming that we are only looking at causes in south and north then

#this is why i removed central because this option means that is separate
x |>
  select(eTT.BG.So, eTT.BG.NN, eTT.BG.FN, REF.GRID) |> 
  filter(REF.GRID != "2 Central") |>
  select(-REF.GRID) |>
  pivot_longer(cols = everything(),
               names_to = "source",
               values_to = "value") |> 
  ggplot(aes(x = source, y = value)) +
  geom_boxplot() +
  labs(
    x = "Source",
    y = "eTT.BG Value",
    title = "Distribution of eTT.BG by Source"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
x<- x %>%
  mutate(
    disp_hour = hour(DT.DISP),
    disp_min  = minute(DT.DISP),
    rush_hour = !is.na(DT.DISP) & (
      
      #morning 7:30pm - 9:30am
      (disp_hour == 7 & disp_min >= 30) |
      (disp_hour == 8) |
      (disp_hour == 9 & disp_min < 30) |
      
      #evening 4:00pm - 6:30pm
      (disp_hour == 16) |
      (disp_hour == 17) |
      (disp_hour == 18 & disp_min < 30)
    ),
    # 1 if rush hour 0 else
    rush_hour_ind = as.integer(rush_hour)  
  ) %>%
  select(-disp_hour, -disp_min)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
time = cbind(x$eTT.BG.NN, x$eTT.BG.FN, x$eTT.BG.So)


x %>%
  select(eTT.BG.NN, eTT.BG.FN, eTT.BG.So) %>%
  pivot_longer(cols = everything(),
               names_to = "source",
               values_to = "time") %>%
  ggplot(aes(x = time, color = source)) +
  geom_density(linewidth = 1.2) +
  labs(
    x = "Time",
    y = "Density",
    title = "Distribution of eTT.BG Values (All Sources)"
  ) +
  theme_minimal()
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
conflicts <- x %>%
  filter(DT.AVAILABLE > DT.DISP) %>%
  mutate(
    conflict_minutes = as.numeric(difftime(DT.AVAILABLE, DT.DISP, units = "mins"))
  )

ggplot(conflicts, aes(x = conflict_minutes)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Conflict Duration Distribution",
       x = "Conflict Duration (minutes)",
       y = "Count") +
  theme_minimal()
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}



# Assume conflicts already has DT.DISP and DT.AVAILABLE
intervals <- IRanges(start = as.numeric(conflicts$DT.DISP),
                     end = as.numeric(conflicts$DT.AVAILABLE))

# Count how many intervals overlap at each start
overlap_counts <- countOverlaps(intervals, intervals)

# Filter rows where there are at least 3 overlapping intervals
conflicts_3plus <- conflicts[overlap_counts >= 3, ]

conflicts <- conflicts %>%
  mutate(overlap_count = overlap_counts,
         overlap_level = ifelse(overlap_count >= 3, "3+ conflicts", "1-2 conflicts"))

ggplot(conflicts, aes(x = conflict_minutes)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Conflict Duration Distribution",
       x = "Conflict Duration (minutes)",
       y = "Count") +
  theme_minimal()
```


### Linear Mixed Model

```{r, echo = FALSE, message = FALSE, warning = FALSE}
formula_latex <- "EstTravelTime ~ Scenario + (1 | CallID)"

# Fit the model
m <- lmer(EstTravelTime ~ Scenario + (1 | CallID), data = x_expanded)

# Extract fixed effects, exclude intercept
coefs <- broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(Significant = !(conf.low <= 0 & conf.high >= 0))

# Plot
ggplot(coefs, aes(x = term, y = estimate, color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  scale_color_manual(values = c("grey60", "steelblue")) +
  coord_flip() +
  labs(x = "", y = "Coefficient estimate",
       title = "Fixed effects (significant ones highlighted)") +
  theme_minimal()
```

## Modeling 

| **Scenarios**  | **S0 (Current)** | **S1** | **S2** | **S3** | **S4** |
|:---------------|:----------------:|:------:|:------:|:------:|:------:|
| **Far North**  |        0         |   0    |   1    |   0    |   1    |
| **Near North** |        0         |   1    |   0    |   1    |   0    |
| **Central**    |        3         |   3    |   3    |   2    |   2    |
| **South**      |        1         |   0    |   0    |   1    |   1    |



```{r, echo = FALSE, message = FALSE, warning = FALSE}
x_filtered <- x_expanded %>%
  group_by(CallID) %>%
  filter(var(EstTravelTime, na.rm = TRUE) > 0) %>%
  ungroup()

mz_filtered <- lmer(
  EstTravelTime ~ Scenario + Distance + rush_hour_ind + (1 | CallID),
  data = x_filtered
)

mz_log_filtered <- lmer(
  log(EstTravelTime) ~ Scenario + Distance + rush_hour_ind + (1 | CallID),
  data = x_filtered
)
```

```{r}


tab_model(
  mz_filtered,
  show.re.var = TRUE,
  show.icc = TRUE,
  show.se = TRUE,
  transform = NULL,
  title = "Linear Mixed Model for EstTravelTime",
  file = "model_table.tex"   # <<--- THIS is the key
)

```



```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Extract residuals and fitted values
resid_filtered <- resid(mz_filtered)
fitted_filtered <- fitted(mz_filtered)

# Residuals vs Fitted
ggplot(data.frame(fitted = fitted_filtered, residuals = resid_filtered),
       aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted (No-Change Calls Removed)",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# QQ plot
ggplot(data.frame(residuals = resid_filtered), aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot of Residuals (No-Change Calls Removed)",
       x = "Theoretical Quantiles", y = "Sample Residuals") +
  theme_minimal()
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
m_varident <- lme(
  EstTravelTime ~ Scenario + Distance + rush_hour_ind,
  random = ~1 | CallID,
  weights = varIdent(form = ~1 | Scenario),
  data = x_filtered,
  method = "REML"
)
# Extract normalized residuals (Pearson) and fitted values
resid_vi <- resid(m_varident, type = "pearson")
fitted_vi <- fitted(m_varident)

ggplot(data.frame(fitted = fitted_vi, residuals = resid_vi),
       aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted (varIdent by Scenario)",
       x = "Fitted Values", y = "Pearson Residuals") +
  theme_minimal()

```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# QQ plot
ggplot(data.frame(residuals = resid_vi), aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot of Residuals (varIdent Model)",
       x = "Theoretical Quantiles", y = "Sample Residuals") +
  theme_minimal()
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Histogram
ggplot(data.frame(residuals = resid_vi), aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Histogram of Residuals (varIdent Model)",
       x = "Pearson Residuals", y = "Count") +
  theme_minimal()
```

